---
title: "project_2"
author: Tejas Kolpek (50559893), Rahul Mannadiar (50560682), Farahath Tabassum (50559788),
  Prathamesh Ravindra Varhadpande (50559586)
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align="center")
```

# Oil Price Time Series Forecasting

## Introduction:

The oil price data from the oil.csv file, which has missing values, will be 
analyzed for this project. In order to investigate trends and seasonality, the 
tasks include plotting the initial time series with missing values, imputing 
missing values, and plotting the time series after imputing the missing values. 
Additionally, we study Holt-Winters and ETS models. We then apply appropriate 
models for our data based on the imputed time series, keeping an eye on trend 
and seasonality. After that, we run the models and assess their suitability. 
Lastly, we evaluate the model's performance using metrics such as the AIC, BIC, 
and Ljung-Box test. Then, we compare the models' performance using RMSE to 
determine which model performs best.


## Step 1. Loading Required Libraries and Data

```{r}
# Load required libraries
suppressWarnings({
library(ggplot2)
library(dplyr)
library(forecast)
library(imputeTS)
})

# Load the dataset
oil_data <- read.csv("D:/fall 2024/statistical learning/oil.csv")
head(oil_data)
```

## Step 2. Plotting the Original Time Series (with missing values)

The date column in the dataset was converted to the Date type to ensure proper 
handling of temporal data. This step is crucial for accurate visualization and 
analysis of time series data. Subsequently, the original time series of oil 
prices was plotted using the ggplot2 library, with the date column on the x-axis 
and the oil price (dcoilwtico) on the y-axis. The plot highlights the trends, 
seasonal patterns, and potential missing values in the raw data, providing an 
initial overview of the dataset's structure and behavior.

```{r}
# Convert date column to Date type
oil_data$date <- as.Date(oil_data$date)

# Plot the time series
ggplot(oil_data, aes(x = date, y = dcoilwtico)) +
  geom_line(color = "blue") +
  labs(title = "Original Time Series", x = "Date", y = "Oil Price") +
  theme_minimal()
```

Trend:

There is a clear downward trend around 2015 and stabilization afterward. 
However, this does not suggest that the seasonality scales with the trend.

Seasonality:

It appears that the seasonal fluctuations do not increase or decrease 
proportionally with the level of the trend. The amplitude of the fluctuations 
seems fairly consistent across the time series, despite the significant drop in 
oil prices around 2015.


## Step 3. Handling Missing Data

We impute the missing data using linear interpolation. This method ensures that 
the imputed values are smooth and realistic.

```{r}
# Impute missing values using linear interpolation
oil_data$dcoilwtico <- na_interpolation(oil_data$dcoilwtico, option = "linear")

# Verify that missing values are handled
sum(is.na(oil_data$dcoilwtico))
```

## Step 4. 
## (a) Plotting the Time Series After Imputation (without missing values)

After imputing the missing data, we plot the time series again to observe if 
there are trends or seasonality.

```{r}
# Plot the imputed time series
ggplot(oil_data, aes(x = date, y = dcoilwtico)) +
  geom_line(color = "green") +
  labs(title = "Time Series with Imputed Data", x = "Date", y = "Oil Price") +
  theme_minimal()
# Convert data to time series
oil_ts <- ts(oil_data$dcoilwtico, start = c(2013, 1), frequency = 365)
```
The trend and seasonality remained same even after imputing the missing values
in the original time series.

Trend:

There is a clear downward trend around 2015 and stabilization afterward. 
However, this does not suggest that the seasonality scales with the trend.

Seasonality:

It appears that the seasonal fluctuations do not increase or decrease 
proportionally with the level of the trend. The amplitude of the fluctuations 
seems fairly consistent across the time series, despite the significant drop in 
oil prices around 2015.

## Step 4. 
## (b) Understanding Trend, Seasonality and Residual Components of Time Series

```{r}
# Decompose the time series into trend, seasonal, and residual components
decomposed_ts <- decompose(oil_ts, type = "additive")  # Since we have constant seasonal variation

# Plot the decomposed components (no 'main' argument)
plot(decomposed_ts)

# Add separate titles to each component manually (optional)
# Plot trend separately for clearer visualization
plot(decomposed_ts$trend, 
     main = "Trend Component of Oil Prices", 
     ylab = "Trend", 
     xlab = "Time", 
     col = "blue", 
     lwd = 2)

# Similarly, you can manually plot seasonal and random components if needed:
plot(decomposed_ts$seasonal, 
     main = "Seasonal Component of Oil Prices", 
     ylab = "Seasonality", 
     xlab = "Time", 
     col = "green", 
     lwd = 2)

plot(decomposed_ts$random, 
     main = "Residual Component of Oil Prices", 
     ylab = "Residuals", 
     xlab = "Time", 
     col = "red", 
     lwd = 2)

```
Analysis:

Trend Component:
The trend component of oil prices shows a significant and consistent decline 
from 2013 to 2015, indicating a downward movement in overall oil price levels 
during this period. After hitting a low point around mid-2015, the trend 
stabilizes and shows slight signs of recovery toward the end of 2015 and into 
early 2016.

Seasonal Component:
The seasonal component exhibits repetitive cyclical patterns across the years, 
suggesting a regular fluctuation in oil prices due to seasonal factors. Peaks 
and troughs recur at similar intervals, possibly linked to changes in energy 
demand based on weather conditions, holiday periods, or other predictable 
events. The amplitude of the seasonal variations remains fairly consistent 
throughout the analyzed period.

Residual Component:

The residual component captures the irregular fluctuations in oil prices that 
cannot be explained by the trend or seasonal components. These residuals display 
high volatility, with periods of sharp increases and decreases.

## Stationarity Check

```{r}
library(tseries)

# Perform the Augmented Dickey-Fuller Test
adf_test <- adf.test(oil_ts)

# Print the results
print(adf_test)

# Interpretation
if (adf_test$p.value <= 0.05) {
  cat("Conclusion: The time series is stationary (reject null hypothesis of unit root).")
} else {
  cat("Conclusion: The time series is not stationary (fail to reject null hypothesis of unit root).")
}
```

Based on the ADF Test we can see the time series is not stationary since the 
p-value is greater than 0.05.

## Step 5. ETS models and Holt-Winter models

ETS Models:

ETS models are a framework for exponential smoothing methods that decompose a 
time series into three components: Error (E), Trend (T), and Seasonality (S). 
These components can each be additive (A), multiplicative (M), or 
non-existent (N). The ETS model is selected based on the best fit to the data 
using criteria like AIC. Additive components imply constant change (linear), 
while multiplicative components account for proportional change. ETS models 
assume errors are independent and identically distributed, and they provide a 
probabilistic foundation for forecasting with uncertainty estimates. In R, ETS 
models can be implemented using the ets() function, which automatically selects 
the best model or allows manual specification.

Holt-Winters Models:

The Holt-Winters model is a specific type of exponential smoothing used for 
time series with trend and seasonality. It comes in two main variations: 
additive (for constant seasonality) and multiplicative (for proportional 
seasonality). The model extends simple exponential smoothing by adding equations 
to smooth the level, trend, and seasonality components. It’s effective for data 
with clear seasonal patterns and assumes the seasonality repeats over a fixed 
period. In R, the Holt-Winters method is implemented using the HoltWinters() 
function, and it works best on stationary data, often requiring preprocessing 
like detrending or deseasonalizing for optimal results.

## Step 6. Selecting suitable models based on data

Based on the analysis of time series data, the ETS model (Error, Trend, Seasonal) 
with seasonal components (e.g., ETS(A,N,M) or ETS(M,A,M)) is a strong option if 
the data exhibits seasonality, as it automatically adjusts for seasonality and 
trends. The Holt-Winters model with multiplicative seasonality is another 
suitable choice, particularly if the seasonal variations are proportional to the 
level of the series, as it performed well in terms of residual autocorrelation. 
Finally, if the data shows complex seasonal patterns and trends, a SARIMA model 
with STLF (Seasonal and Trend decomposition using Loess) can provide a more 
robust solution by automatically handling intricate seasonality and trend 
decomposition. Testing and comparing these models based on forecast accuracy and 
residual analysis will help determine the most suitable model for your data.

## Step 7. Implementing suitable models 

## (a) ETS Model

```{r}
library(forecast)
suppressWarnings({
# Fit an ETS model to a time series
ets_model <- ets(oil_ts)

# Summary of the model
summary(ets_model)

# Print ETS Model  
print(ets_model)

# Plot the forecast
forecast_ets <- forecast(ets_model, h = 12)  # Forecast for the next 12 periods
plot(forecast_ets)
})
```
The output indicates that the ETS(A,N,N) model was selected for the time series 
oil_ts, which means the model uses additive error (A), no trend (N), and no 
seasonality (N). The smoothing parameter alpha is 0.9683, showing a high weight 
given to recent observations. The initial level (l) of the series is 93.1395, 
and the residual standard deviation (sigma) is 1.1771. The model's fit is 
quantified by several metrics: the AIC (9055.133), AICc (9055.153), and 
BIC (9070.448), which are used to assess model performance, with lower values 
indicating better fits. The training set error measures show a mean error (ME) 
close to zero (-0.0389), suggesting minimal bias. RMSE (1.1762) and MAE (0.8951) 
reflect the average deviations from the true values, while MAPE (1.5483%) 
suggests a small percentage error in forecasts. The ACF1 value (-0.00075) 
indicates no significant autocorrelation in the residuals. However, the warning 
about seasonality being ignored (due to a frequency greater than 24) suggests 
that seasonality is not being modeled, and the stlf() function may be needed for 
seasonal adjustments.


## Step 5.(b) Holt-Winters Model

```{r}
# Holt-Winters Additive Model
hw_additive <- HoltWinters(oil_ts, seasonal = "additive")

# Forecast for Additive Model
hw_additive_forecast <- forecast(hw_additive, h = 12)
plot(hw_additive_forecast)

# Holt-Winters Multiplicative Model
hw_multiplicative <- HoltWinters(oil_ts, seasonal = "multiplicative")

# Forecast for Multiplicative Model
hw_multiplicative_forecast <- forecast(hw_multiplicative, h = 12)
plot(hw_multiplicative_forecast)

# Check residuals for the models
checkresiduals(hw_additive_forecast)
checkresiduals(hw_multiplicative_forecast)
```
The Ljung-Box test for the residuals from the Holt-Winters additive model 
resulted in a p-value of 0.09457, which is slightly above the typical 
significance threshold of 0.05. While this suggests some evidence of 
autocorrelation in the residuals, the p-value is close to the threshold, 
indicating marginal or weak autocorrelation. Therefore, while the additive model 
may be generally capturing the time series dynamics, the presence of some 
residual correlation suggests that the model could potentially be improved, 
possibly by refining the seasonal component or adding more complexity.

The Ljung-Box test for the residuals from the Holt-Winters multiplicative model 
yielded a p-value of 0.8647, which is well above the 0.05 threshold, indicating 
no significant autocorrelation in the residuals. This suggests that the 
multiplicative model adequately captures the seasonality and underlying 
structure of the data, with residuals behaving like white noise. The lack of 
significant autocorrelation supports the effectiveness of the multiplicative 
model in fitting the time series, indicating that it might be a better choice 
in comparison to the additive model based on these residuals.

## (c) SARIMA Model

```{r}
# Auto ARIMA with Seasonal Component
sarima_model <- auto.arima(oil_ts, seasonal = TRUE)

# Forecast
sarima_forecast <- forecast(sarima_model, h = 12)
plot(sarima_forecast)

# Check residuals
checkresiduals(sarima_forecast)
```
The Ljung-Box test for the residuals from the ARIMA(0,1,0) model yielded a 
p-value of 0.197, which is well above the typical significance threshold of 
0.05. This indicates that there is no significant autocorrelation in the 
residuals, suggesting that the model adequately captures the underlying patterns 
in the data. The residuals appear to behave like white noise, meaning the model 
has effectively accounted for the time series dynamics, and no further 
improvements in the model specification are needed.

Additionally, the Auto ARIMA model with a seasonal component was fitted to the 
data, followed by forecasting for the next 12 periods and a residual check using 
checkresiduals(). This further confirms that the SARIMA model adequately handles 
both seasonality and trend in the data, with the residuals showing no 
significant autocorrelation.

## (d) SARIMA Model with STLF Function

```{r}
# STL Decomposition and Forecast
stl_forecast <- stlf(oil_ts, h = 12, method = "arima")  # Use ARIMA for forecasting
plot(stl_forecast)

# Residual diagnostics
checkresiduals(stl_forecast)
```
The Ljung-Box test for the residuals from the STL decomposition combined with 
ARIMA(0,1,0) model yielded a p-value of 0.0002194, which is highly significant 
and below the typical threshold of 0.05. This indicates the presence of 
significant autocorrelation in the residuals, suggesting that the model has not 
fully captured all the underlying patterns in the data. Therefore, the residuals 
are not behaving like white noise, and the model might need further refinement, 
such as adjustments to the ARIMA parameters or additional seasonal components.

Following this, the STL decomposition with ARIMA method was applied to forecast 
the next 12 periods, and residual diagnostics were performed using 
checkresiduals() to evaluate the adequacy of the model fit.

## Step 8. Comparing the models based on AIC, BIC, Ljung-Box p-value and RMSE

```{r}
suppressWarnings({
# Fit ETS model
ets_model <- ets(oil_ts)

# Calculate metrics
rmse_ets <- sqrt(mean((oil_ts - fitted(ets_model))^2, na.rm = TRUE))  # RMSE
aic_ets <- AIC(ets_model)  # AIC
bic_ets <- BIC(ets_model)  # BIC

# Ljung-Box test
ljung_ets <- Box.test(residuals(ets_model), lag = 20, type = "Ljung-Box")

# Print results
cat("ETS Model:\n")
cat("RMSE:", rmse_ets, "\nAIC:", aic_ets, "\nBIC:", bic_ets, "\nLjung-Box p-value:", ljung_ets$p.value, "\n\n")
})


# Fit Holt-Winters Additive Model
hw_model <- HoltWinters(oil_ts, seasonal = "additive")

# Forecast
hw_forecast <- forecast(hw_model, h = 12)

# Calculate RMSE
rmse_hw <- sqrt(mean((oil_ts - fitted(hw_model))^2, na.rm = TRUE))  # RMSE

# For AIC and BIC
n_hw <- length(oil_ts)
rss_hw <- sum((residuals(hw_model))^2, na.rm = TRUE)  # Residual sum of squares
aic_hw <- n_hw * log(rss_hw / n_hw) + 2 * 3  # 3 params: alpha, beta, gamma
bic_hw <- n_hw * log(rss_hw / n_hw) + log(n_hw) * 3

# Ljung-Box test
ljung_hw <- Box.test(residuals(hw_model), lag = 20, type = "Ljung-Box")

# Print results
cat("Holt-Winters Additive Model:\n")
cat("RMSE:", rmse_hw, "\nAIC:", aic_hw, "\nBIC:", bic_hw, "\nLjung-Box p-value:", ljung_hw$p.value, "\n\n")



# Fit Holt-Winters Multiplicative Model
hw_model_multiplicative <- HoltWinters(oil_ts, seasonal = "multiplicative")

# Forecast
hw_forecast_multiplicative <- forecast(hw_model_multiplicative, h = 12)

# Calculate RMSE
rmse_hw_multiplicative <- sqrt(mean((oil_ts - fitted(hw_model_multiplicative))^2, na.rm = TRUE))  # RMSE

# For AIC and BIC
n_hw_mul <- length(oil_ts)
rss_hw_mul <- sum((residuals(hw_model_multiplicative))^2, na.rm = TRUE)  # Residual sum of squares
aic_hw_mul <- n_hw_mul * log(rss_hw_mul / n_hw_mul) + 2 * 3  # 3 params: alpha, beta, gamma
bic_hw_mul <- n_hw_mul * log(rss_hw_mul / n_hw_mul) + log(n_hw_mul) * 3

# Ljung-Box test
ljung_hw_mul <- Box.test(residuals(hw_model_multiplicative), lag = 20, type = "Ljung-Box")

# Print results
cat("Holt-Winters Multiplicative Model:\n")
cat("RMSE:", rmse_hw_multiplicative, "\nAIC:", aic_hw_mul, "\nBIC:", bic_hw_mul, "\nLjung-Box p-value:", ljung_hw_mul$p.value, "\n\n")



# Fit Seasonal ARIMA Model
sarima_model <- auto.arima(oil_ts, seasonal = TRUE)

# Calculate metrics
rmse_sarima <- sqrt(mean((oil_ts - fitted(sarima_model))^2, na.rm = TRUE))  # RMSE
aic_sarima <- AIC(sarima_model)  # AIC
bic_sarima <- BIC(sarima_model)  # BIC

# Ljung-Box test
ljung_sarima <- Box.test(residuals(sarima_model), lag = 20, type = "Ljung-Box")

# Print results
cat("Seasonal ARIMA Model:\n")
cat("RMSE:", rmse_sarima, "\nAIC:", aic_sarima, "\nBIC:", bic_sarima, "\nLjung-Box p-value:", ljung_sarima$p.value, "\n\n")



# STL decomposition with ARIMA
stl_forecast <- stlf(oil_ts, h = 12, method = "arima")

# Calculate RMSE
fitted_stl <- oil_ts - residuals(stl_forecast$model)
rmse_stl <- sqrt(mean((oil_ts - fitted_stl)^2, na.rm = TRUE))  # RMSE

# AIC and BIC
aic_stl <- AIC(stl_forecast$model)  # AIC
bic_stl <- BIC(stl_forecast$model)  # BIC

# Ljung-Box test
ljung_stl <- Box.test(residuals(stl_forecast$model), lag = 20, type = "Ljung-Box")

# Print results
cat("STL with ARIMA Model:\n")
cat("RMSE:", rmse_stl, "\nAIC:", aic_stl, "\nBIC:", bic_stl, "\nLjung-Box p-value:", ljung_stl$p.value, "\n\n")
```
Based on the provided metrics, the STL with ARIMA Model has the lowest RMSE of 
0.9799459, which indicates that it provides the best fit in terms of forecasting 
accuracy among all the models. This is followed closely by the ETS Model 
(RMSE = 1.176176), which also performs well in terms of forecast accuracy. 
In contrast, the Holt-Winters Additive Model (both versions) shows significantly
higher RMSE values (40.72244 and 40.62907), indicating poorer accuracy compared 
to the other models.

Additionally, the Ljung-Box p-value values suggest that residuals from the STL 
with ARIMA Model (p-value = 0.5681022) and Seasonal ARIMA Model 
(p-value = 0.8994111) exhibit no significant autocorrelation, meaning these 
models adequately capture the underlying patterns in the data.

Thus, the STL with ARIMA Model is the preferred model based on the lowest RMSE, 
along with favorable residual diagnostics.
